Hidden Multiple Comparisons Increase Forensic Error Rates

### 2024-04-01 PNAS/reviews/response.qmd

---
title: "Reviewer Response"
author: "Susan Vanderplas, Heike Hofmann, Alicia Carriquiry"
format: pdf
---

# General Comments

We thank both reviewers for their insightful comments. While we respond point-by-point below, the reviews of this paper are both careful, considerate, and helpful, and will serve to improve this paper.

# Reviewer 1

- **Points 1 and 2**: Existing alternative solutions to multiple comparisons problems and existing psychological literature that speaks to this kind of comparison task.

> We appreciate the reviewer's helpful suggestions about additional situations which may be similar and/or which may help to resolve the multiple comparison problem. One extremely unique factor of wire cuts that is not true of duct tape fits, bullets, or cartridge cases is that there is not any external structure which can assist with the multiple comparison problem. In duct tape, the tape has edges (and internal fibers) that provide structure. In bullets, there are lands and grooves which provide some structure and a way to sequence the lands according to the physical bullet orientation; this reduces a $n^2$ comparison problem to an $n$ comparison problem, where $n$ is the number of lands (though of course, database searches make even this a challenge). In cartridge cases, the extractor marks are of limited width and comparable size to the mechanism itself. In contrast, with wire cuts, there is a blade cut and a wire, and the wire could have been cut at any point along the blade; in most cases there is not any structure beyond the striation marks themselves. Thus, wire cuts serve as a uniquely problematic example within the wider issue of multiple comparisons in forensics. The Prusinowski paper is a good paper, but the holistic judgement process demonstrated in e.g. Fig 7 of the manuscript is predicated on edge alignment as a first step; no physical analogue to this exists for wire cuts. While the Prusinowski paper's method is one way to avoid cross-correlation, it inherently relies on an assumption which is not true in most striation marks: that the length of the striated region is the same for the evidence as for the comparison sample.

> Similarly, the psychological concepts relating to holistic judgements are difficult to apply to wire cuts and even bullets, because examination of the striations visually precludes examination of the entire entity, either due to the length (e.g. the blade cut is many times longer than the wire diameter) or due to curvature (in the case of bullets). While it is possible that a talented cognitive psychologist could provide an improved examination procedure that would systematize this sort of comparison, we aren't aware of any such procedures implemented in labs at the moment that are aimed at addressing this problem; in addition, as we are limited to 3 pages, it seems quite impossible to do the topic any justice. We would be most interested to collaborate with Reviewer 1 on an exloration of the zoomed-in vs. holistic cognition of striated mark comparisons, if they are a cognitive psychologist, as two of the three authors of this paper are statisticians who occasionally dabble in psychology-adjacent disciplines such as perception of statistical graphics.

### 2024-04-01 PNAS/index.Rnw
A few minor tweaks + reviewer response letter

\dropcap{I}n forensic evaluations, a single conclusion often relies on  many comparisons, either implicitly or explicitly.
Multiple comparisons arise persistently when developing statistical methods to address scientific problems \citep{benjaminiControllingFalseDiscovery1995}, and greatly increase the probability of false discoveries.
Now that vast databases and efficient algorithms are routinely used in forensic evaluations to propose matches to crime scene items, the problem of close non-matches~\citep{pcast} due to multiple comparisons becomes critically important.
This often ignored issue increases the false positive rate, and can contribute to the erosion of public trust in the justice system.
The multiple comparison problem is not new: it has been raised in the past with regard to DNA~\citep{thompson2003} and latent print evaluations~ \citep{koehler2021}. One of the root causes~\citep{fine2006} leading to the wrongful accusation of Brandon Mayfield in the 2004 Madrid train bombing case was that the large size of the IAFIS database used to search for similar prints made it possible to locate `unusually' close non-matches.
% The probability of finding  close non-matches for an item of evidence is proportional to the size of the database available for searching; close non-matches occur more frequently as databases increase in size.
As database size increases, so does the probability of finding a close non-match.

Compounding this issue, the use of algorithms also results in a large number of comparisons that are not obvious to the user.
For example, the cross-correlation function~\citep{vorburgerApplicationsCrosscorrelationFunctions2011} was one of the first measures proposed for quantifying the similarity between two patterns in response to the 2009 NRC report~\citep{nas2009}, and continues to be used in many pattern searching algorithms to find the best alignment between two images and to quantify their overall similarity.
Finding the best alignment often consists in sliding one surface across the whole length (for one-dimensional patterns, such as striations) or area (for two dimensional sources, such as impression marks) of the other item while keeping track of the value of a similarity measure.
This mirrors the forensic examination process: the examiner visually rotates and shifts items under a comparison microscope to align two surfaces.
In order to avoid false accusations and the corresponding impact on public perception of forensics, we must address the problem of multiple comparisons and control their effect on false discovery rates.

>>>>

\dropcap{I}n forensic evaluations, a single conclusion often relies on  many comparisons, either implicitly or explicitly.
Multiple comparisons arise persistently when developing statistical methods to address scientific problems \citep{benjaminiControllingFalseDiscovery1995}, and greatly increase the probability of false discoveries.
Now that vast databases and efficient algorithms are routinely used in forensic evaluations to propose matches to crime scene items, the problem of close non-matches~\citep{pcast} due to multiple comparisons becomes critically important.
This often ignored issue increases the false discovery rate, and can contribute to the erosion of public trust in the justice system through conviction of innocent individuals.
The multiple comparison problem is not new: it has been raised in the past with regard to DNA~\citep{thompson2003} and latent print evaluations~ \citep{koehler2021}. One of the root causes~\citep{fine2006} leading to the wrongful accusation of Brandon Mayfield in the 2004 Madrid train bombing case was that the large size of the IAFIS database used to search for similar prints made it possible to locate `unusually' close non-matches.
As database size increases, so does the probability of finding a close non-match.

Compounding this issue, the use of algorithms also results in a large number of comparisons that are not obvious to the user.
For example, the cross-correlation function~\citep{vorburgerApplicationsCrosscorrelationFunctions2011} was one of the first measures proposed for quantifying the similarity between two patterns in response to the 2009 NRC report~\citep{nas2009}, and continues to be used in many pattern searching algorithms to find the best alignment between two images and to quantify their overall similarity.
Finding the best alignment often consists in sliding one surface across the whole length (for one-dimensional patterns, such as striations) or area (for two dimensional sources, such as impression marks) of the other item while keeping track of the value of a similarity measure.
This mirrors the forensic examination process: the examiner visually rotates and shifts items under a comparison microscope to align two surfaces.
In order to avoid false accusations and the corresponding impact on public perception of forensics, we must address the problem of multiple comparisons in database and alignment searches and control their effect on false discovery rates.
